---
title: "Plotting Assignment #7"
subtitle: "Spatial Plots"
author: "Haley Brumberger"
date: "`r format(Sys.Date(), '%d %b %Y')`"
number-sections: true
number-offset: 0
toc: true
toc-depth: 2
fig-width: 6
fig-height: 4
df-print: kable
embed-resources: true
format: 
  html: 
    highlight-style: dracula
    code-tools: true
    code-fold: show
    code-summary: "Show the code"
    code-line-numbers: true
    code-link: true # only works if code-line-numbers: false
    toc-float: true
  docx: 
    toc-title: "Table of contents"
    fig-dpi: 600
    execute:
      echo: false
knitr: 
  opts_chunk: 
    fig.path: "plots/"
    fig.keep: "all"
    dev: ['png', 'pdf']
    dev.args: 
      pdf: 
        encoding: 'WinAnsi'
        useDingbats: false
bibliography: references.bib # if using references, need a .bib file
csl: gpl.csl # find your preferred citation style at https://www.zotero.org/styles
css: style.css # optional html styling
crossref:
  fig-prefix: Figure
  tbl-prefix: Table
  ref-hyperlink: true
editor: source
---

# Setup

Using `r R.version.string` , @R, and ggplot2 version `r packageVersion("ggplot2")` from the tidyverse [@tidyverse].

```{r}
#| label: setup
#| message: false
#| warning: false

# load packages
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
library(lubridate)
library(rnaturalearth)
library(ggrepel)
```

# Load Data

```{r}
#| message: false
#| warning: false
# loading well data from parquet in the data folder
wells <- arrow::read_parquet("data/monitoring_wells.parquet") |>
  st_as_sf(coords = c("DecLongVa", "DecLatVa")) |> # setting coordinates (Long, Lat) according to the named columns
  st_set_crs("NAD83") |> # set coordinate system the well data is in
  st_transform("WGS84") # transform coordinate system to WGS84

wells <- wells |> filter(aquifer_name != "NA")

# switch off spherical geometry to allow planar centroid calculations
sf_use_s2(FALSE)

# loading aquifer data
aquifers <- sf::read_sf("data/aquifers/Bedrock Aquifer.shp") |>
  st_transform("WGS84")

# load water level data, add a new column that is water_level.ft to be the depth below the surface as negative numbers from the pre-existing depth column
water_levels <- arrow::read_parquet("data/waterlevels.parquet") |>
  mutate("water_level.ft" = (`Depth to Water Below Land Surface in ft.` * -1))


# load lithology data
lithology <- sf::read_sf("data/CO_lithology/CO_geol_poly.shp") |>
  tidyr::separate(
    "GENERALIZE",
    into = c("major_rock_type", "rock_type_details"), # throws a warning that missing pieces are filled with "NA", these columns are the WATER columns (not rocks)
    sep = ","
  )

# load counties data
counties_co <- maps::map("county", fill = TRUE) |>
  st_as_sf() |>
  tidyr::separate(ID, into = c("state", "county"), sep = ",") |> # split the ID into "state" and "county" which are separated by commas
  st_as_sf() |>
  filter(state == "colorado") |> # filter to only get counties in colorado
  st_transform("WGS84") #  transform into WGS84 coordinate system

# states
states <- ne_states(country = "United States of America") |>
  st_as_sf()

# focus on colorado
co <- states |> filter(name == "Colorado")
co_bbox <- co |> st_bbox()

```

## Exploratory Plots/Data Vis

```{r}
#| warning: false

# # plot wells
# ggplot() + geom_sf(data = wells, map = aes(color = aquifer_name))

# plot aquifers
# ggplot() +
#   geom_sf(data = aquifers, map = aes(color = NAME)) +
#   geom_sf_label(
#     data = aquifers,
#     map = aes(
#       color = NAME,
#       label = NAME,
#       size = ,
#       geometry = st_centroid(geometry)
#     ),
#     show.legend = FALSE
#   )

# # plot water levels
# water_levels |>
#   group_by(aquifer_name) |>
#   summarize(mean_water_level.ft = mean(water_level.ft))

# # plot lithology
# ggplot() + geom_sf(data = lithology, map = aes(fill = major_rock_type))

# # plot colorado counties
# ggplot() +
#   geom_sf(data = counties_co) +
#   geom_sf_text(data = counties_co, map = aes(label = county))
```

# Plots

# Plot 1: Geologic Map of Colorado with Monitoring Wells
@fig-1, a geologic map, shows the lithology of Colorado, illustrated by colors, with the locations of six aquifers and their associated monitoring wells overlain. The igneous and metamorphic rock shown (in the middle in purple and pink) represent the Front Range mountain range, with the Denver Basin westernmost extent nestled right up against the foothills. Water is the least common surface feature compared to all of the rock types. Most of the aquifers are found in igneous and sedimentary bedrock, but this is likely due to the fact that Colorado is mostly igneous and sedimentary bedrock. The High Plains basin has the most number of monitoring wells (n = 126) in Colorado, while the Paradox and San Juan Basins have 1 and 2 wells, respectively, in the state. The High Plains Basin wells are relatively well spaced out, while the Piceance Basin and San Luis Basin wells are closer together overall--and there are fewer in the latter two. On this map, the High Plains basin is the largest (in the real world it covers ~112 million acres), whereas the San Luis Basin is the smallest (~2 million acres in the real world; both numbers from the Bureau of Land Management website). 
```{r}
#| message: false
#| warning: false
#| label: fig-1
#| fig-height: 6
#| fig-width: 10
#| fig-cap: Figure showing Colorado lithology (rock type), extent of six CO aquifers, and locations of wells in each of the aquifers. Colors correspond to different rock types and different basins.

# making a colors vector for the rock types
colors <- c(
  "Igneous" = "#FF6F91",
  "Sedimentary" = "#92DCB7",
  "Metamorphic" = "#A7A7FF",
  "Metamorphic and Sedimentary" = "#7DFF7D",
  "Igneous and Sedimentary" = "#60CCBF",
  "Unconsolidated" = "#FDF43F",
  "Unconsolidated and Sedimentary" = "#FFFF89",
  "Water" = "#99CCFF"
)

# filtering aquifers and wells to remove NA values so that I can plot only the ones with monitoring wells
filtered_aquifers <- aquifers |>
  filter(!is.na(NAME)) |>
  st_filter(wells)

wells <- wells |>
  filter(!is.na(aquifer_name))


# making the plot, geologic map of colorado with monitoring wells/aquifer locations
lithology |>
  ggplot() + # creating the plot frame
  geom_sf(aes(fill = major_rock_type)) + # adding the major rock type as a spatial feature layer, with fill according to the different rock type
  scale_fill_manual(values = colors) + # setting the fill colors to be the ones I determined in the colors vector above
  geom_sf(
    # adding aquifer information/plotting the aquifers
    data = filtered_aquifers, # filtering by those who have wells
    aes(color = NAME), # coloring them by aquifer name
    show.legend = FALSE, # turning the legend off for this geom because otherwise it adds a polygon to the legend entry, when all you really see are points
    alpha = 0.6
  ) +
  geom_sf(
    # adding well information/plotting the wells
    data = wells,
    aes(color = aquifer_name), # coloring wells by aquifer name
    size = 1 # making the points smaller
  ) +
  scale_color_brewer(palette = "Dark2") + # making the points darker so they show up better
  geom_sf_label(
    # adding aquifer labels
    data = filtered_aquifers,
    aes(label = NAME, geometry = st_centroid(geometry)),
    stat = "sf_coordinates",
    color = "black", # color of the outline
    size = 3, # making the text smaller
    alpha = 0.5, # text box opacity
  ) +
  annotation_scale(
    # adding a scale bar
    location = "tl", # top left location
    width_hint = 0.1, # making scale bar narrower so it doesn't take up as much room
    pad_x = unit(0.5, "cm"), # adding space between the top edge of the plot and the scale bar
    pad_y = unit(0.4, "cm") # adding space between the left edge of the plot and the scale bar
  ) +
  annotation_north_arrow(
    # adding a north arrow
    style = north_arrow_orienteering(), # specifying north arrow type/design
    location = "tl", # top left location
    height = unit(0.1, "npc"), #making the north arrow shorter to take up less room
    width = unit(0.05, "npc"), # making the north arrow narrower
    pad_x = unit(0.85, "cm"), # adding space between left edge of plot and north arrow, centering underneath scale bar
    pad_y = unit(0.8, "cm") # adding space between top edge of plot and north arrow, ensuring underneath scale bar
  ) +
  coord_sf(
    # cropping the plot to the edge of the data frame
    xlim = c(co_bbox["xmin"] + 0.31, co_bbox["xmax"] - 0.32),
    ylim = c(co_bbox["ymin"] + 0.21, co_bbox["ymax"] - 0.21)
  ) +
  theme_bw() + # setting a b&w theme to get rid of grey background
  labs(
    # changing labels for x/y axes, title, and legend titles
    x = "Longitude",
    y = "Latitude",
    title = "Geologic Map of Colorado with Monitoring Wells (and Aquifers)",
    fill = "Major Rock Type",
    color = "Aquifer/Wells"
  ) +
  theme(
    # customizing the plot
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), # centering, sizing, and bolding title
    axis.title = element_text(size = 11, face = "bold"), # sizing and bolding axis titles
    legend.title = element_text(size = 10, face = "bold"), # sizing and bolding legend titles
    axis.text = element_text(size = 9), # sizing axis text
    legend.text = element_text(size = 9) # sizing legend text
  ) +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) # setting the order of the legend so that major rock type comes first

```

## Plot 2: Mean Change in Water Levels Over Time
@fig-2 displays mean yearly water level change from 1980-2020 for the six aquifers shown in @fig-1. All of the aquifers experienced both gains and losses over time, but the High Plains aquifer, which is the largest on the map, experiences the most dramatic loss of water level by 2020, losing 23.15 feet of water over 40 years. The Paradox aquifer, meanwhile, gained 9.95 feet of water in those same years; however, it is notable that the Paradox aquifer only has 1 monitoring well, so that data comes from a single well over 40 years. The most dramatic year-to-year fluctations occur in the Piceance Basin, with the water level from 1990 to 1991 increasing by 13.66 feet and the water level from 1993 to 1994 decreasing by 17.31 feet; water level then increases by 12.97 feet between 1994 and 1995. In the Denver Basin, there is a steady decrease in water level between 1980 and 2003, then the aquifer turns a corner and water level increases 7.62 feet between 2006-2007. This steady decrease was likely a function of population growth increasing the demand for water, groundwater pumping, variable precipitation, and drought (@https://doi.org/10.1111/1752-1688.12755). By 2020, the water level in the Denver Basin had increased 4.54 feet--a positive sign for the 2.5+ million people who live in the area. Unfortunately, this year (2026)'s historically low snowpack and winter precipitation totals will likely contribute to further aquifer depletion in the coming year.
```{r}
#| message: false
#| warning: false
#| label: fig-2
#| fig-height: 6
#| fig-width: 10
#| fig-cap: Figure showing change in water level over time in the six Colorado aquifers from Figure 1. The black dashed line represents no change in water level, the point size corresponds to number of wells, and the color corresponds to aquifer name.

# making a new dataframe with mean water levels and the differences between the yearly mean and the first year's mean
water_levels_mean <- water_levels |>
  mutate(year = year(Time)) |> # pulling the year from the Time column
  summarize(
    .by = c(SiteNo, aquifer_name, year), # grouping by aquifer, Site Number and new year column
    annual_water_level.ft = mean(water_level.ft, na.rm = TRUE)
  ) |>
  group_by(SiteNo, aquifer_name) |>
  mutate(
    # add a column calculating the difference between each year's mean and the original mean water level
    initial_water_level.ft = annual_water_level.ft[year == min(year)], # original water level
    change_in_water_level.ft = annual_water_level.ft - initial_water_level.ft # change in water level
  ) |>
  group_by(year, aquifer_name) |>
  summarize(
    mean_change_in_water_level.ft = mean(
      # calculate average change in water level per aquifer per year
      change_in_water_level.ft,
      na.rm = TRUE
    ),
    n_wells = n() # counting how many wells contributed to the overall aquifer average each year
  )

# filtering out NA values
water_levels_mean <- water_levels_mean |>
  filter(!is.na(aquifer_name))

# plotting the plot
water_levels_mean |>
  ggplot() +
  aes(
    # mapping aesthetics
    x = year,
    y = mean_change_in_water_level.ft,
    size = n_wells, # sizing according to how many wells were counted in the average
    color = aquifer_name # coloring by aquifer name to distinguish between them
  ) +
  scale_size_continuous(breaks = c(1, 30, 60, 90, 120, 150)) +
  # scale_size_area(n.breaks = 6, max_size = 6) +
  geom_point() + # adding points
  scale_color_brewer(palette = "Paired") + # using a colorblind friendly palette
  geom_line(
    # adding lines
    inherit.aes = FALSE, # changing the aesthetics so size doesn't carry over to the line width
    data = water_levels_mean, # setting the dataframe to pull data from
    aes(x = year, y = mean_change_in_water_level.ft, color = aquifer_name) # same parameters as original aesethetic but without the size category; tried to make size only a geom_point aesthetic but it said it couldn't "find" the n_wells column in that dataset
  ) +
  geom_hline(
    # adding a dashed horizontal line with a y intercept of 0 to highlight what "no change" would look like
    aes(yintercept = 0, linetype = "No Change (0 ft)"), # putting these inside the aesthetics so they can be mapped to a legend entry
    linewidth = 1, # setting linewidth outside of aes so it does NOT get mapped to a legend entry
  ) +
  scale_linetype_manual(values = c("No Change (0 ft)" = "longdash")) +
  theme_bw() + # black and white theme to get rid of grey background
  labs(
    # changing labels for x/y axes, title, and legend titles
    x = "Year",
    y = "Change in Water Level (ft)",
    title = "Change in Water Level per Year\nin Six Colorado Aquifers",
    color = "Aquifer Name",
    size = "Number of Wells",
    linetype = "Reference"
  ) +
  theme(
    # customizing the plot
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), # centering, sizing, and bolding title
    axis.title = element_text(size = 11, face = "bold"), # sizing and bolding axis titles
    legend.title = element_text(size = 10, face = "bold"), # sizing and bolding legend titles
    axis.text = element_text(size = 9), # sizing axis text
    legend.text = element_text(size = 9) # sizing legend text
  )
```

## Plot 3: Water Level Change of Individual Wells
@fig-3 shows the same six Colorado aquifers as in @fig-1 and @fig-2, but this time also shows the 40-year water level difference in each individual monitoring wells across the state. The forty year water level difference ranges from -67.67 feet to +71.71 feet across different wells--with one outlier in the Piceance Basin at +244.30 feet--and the average forty year difference across all wells is -7.77 feet. Individual aquifer average difference per year ranges from -17.33 to +14.21 feet: 
                              High Plains: -17.33;
                              San Luis: -6.51;
                              Denver: +2.25;
                              Paradox: +7.01;
                              San Juan: -6.71;
                              Piceance: +14.21.
Overall, drought conditions in Colorado fluctuate, but seem to be becoming more serious over time. Additionally, this year's exceptionally low snowpack and limited precipitation will probably have lasting impacts on the state's water supply, possibly for years to come. 

```{r}
#| message: false
#| warning: false
#| label: fig-3
#| fig-height: 6
#| fig-width: 10
#| fig-cap: Figure showing six Colorado aquifers and the change in well water levels over the last forty years in those aquifers; the aquifers are separated by color while the forty year difference (in feet) in well water level is a greyscale from white (lowest, a.k.a highest decrease in water level) to black (highest increase in water level). The red points represent wells with no data.

# new df with water level averages in the 1980s per well
water_level_1980s <- water_levels |>
  mutate(year = year(Time)) |> # pulling out year from Time column
  filter(year >= 1980 & year < 1990) |> # filtering between 1980-1990
  summarize(
    # summarizing by well and averaging water level
    .by = c(SiteNo),
    water_level_1980s = mean(water_level.ft, na.rm = TRUE)
  )


# new df with water level averages in the 2010s per well
water_level_2010s <- water_levels |>
  mutate(year = year(Time)) |> # pulling out year from Time column
  filter(year >= 2010 & year < 2020) |> # filtering between 2010-2020
  summarize(
    # summarizing by well and averaging water level
    .by = SiteNo,
    water_level_2010s = mean(water_level.ft, na.rm = TRUE)
  )

# making a new dataframe to be able to put only corresponding aquifer data back into wells_w_water_level_diffs df
well_info <- water_levels |>
  select(SiteNo, aquifer_name) |> # keep SiteNo and aquifer_name
  distinct() # only the unique combos to remove duplicates and ensure smooth joining

wells_info <- wells |> # making new df with just SiteNo and geometry of wells for plot
  select(SiteNo, geometry)

# adding aquifer data back in, joined by SiteNo
wells_w_water_level_diffs <- left_join(
  water_level_1980s,
  water_level_2010s,
  by = "SiteNo"
) |>
  left_join(well_info, by = "SiteNo") |>
  left_join(wells_info, by = "SiteNo") # adding the geometry data back in from the wells_info df which came from the well df

wells_w_water_level_diffs <- wells_w_water_level_diffs |>
  mutate(
    forty_yr_diff = water_level_2010s - water_level_1980s
  )

# calculating average water level gain/drop over forty year
averages <- wells_w_water_level_diffs |>
  summarize(
    .by = "aquifer_name",
    mean_per_aquifer = mean(forty_yr_diff)
  )


# plotting the plot
states |>
  filter(name == "Colorado") |> # filtering states to only Colorado
  ggplot() + # making the plot frame
  geom_sf(
    # adding counties in Colorado geometries; these are the polygons
    data = counties_co,
    aes(geometry = geom),
    color = "black",
    fill = "white",
  ) +
  geom_sf_text(
    # adding county names in Colorado that correspond to the polygons in the layer above
    data = counties_co,
    aes(label = county), # setting the label to be county
    size = 2, # making the text smaller
    alpha = 0.5 # setting the text to be a little transparent so they're easier to ignore/the plot is less busy
  ) +
  geom_sf(
    data = filtered_aquifers, # plotting only aquifers with monitoring wells
    aes(geometry = geometry, fill = NAME), # setting the geometry to be the geometry column, filling (color fill) by name
    alpha = 0.75 # setting opacity
  ) +
  geom_sf(
    # plotting differences in well water level from 1980-2020
    data = wells_w_water_level_diffs, # setting dataframe to pull data from
    aes(geometry = geometry, color = forty_yr_diff) # setting geometry to be the geometry column, and the color to be the difference between 1980s average and 2010s average
  ) +
  scale_color_gradient(
    # customizing the color gradient
    low = "#ffffff", # making the low white
    high = "#000000", # making the high black
    limits = c(-50, 50), # setting the lower and upper limits of the scale
    #midpoint = 0, # not necessary if not doing gradient2 since it is a continuous scale
    na.value = "red" # setting the well locations with no data to be red points
  ) +
  geom_sf_label(
    # adding aquifer labels
    data = filtered_aquifers, # setting the dataframe to pull data frame
    aes(label = NAME, fill = NAME), # labeling the aquifers and filling the label with a color according to its name
    color = "black", # color of the outline
    size = 3, # making the text smaller
    alpha = 0.75, # text box opacity
    show.legend = FALSE, # removing legend from this so that the aquifer only has one entry
  ) +
  scale_fill_brewer(palette = "Set2") + # using colorblind friendly palette
  guides(color = guide_colorsteps(show.limits = TRUE)) + # showing upper and lower limit of scale
  annotation_scale(
    # adding a scale bar
    location = "tl", # top left location
    width_hint = 0.1, # making scale bar narrower so it doesn't take up as much room
    pad_x = unit(0.5, "cm"), # adding space between the top edge of the plot and the scale bar
    pad_y = unit(0.4, "cm") # adding space between the left edge of the plot and the scale bar
  ) +
  annotation_north_arrow(
    # adding a north arrow
    style = north_arrow_orienteering(), # specifying north arrow type/design
    location = "tl", # top left location
    height = unit(0.07, "npc"), #making the north arrow shorter to take up less room
    width = unit(0.035, "npc"), # making the north arrow narrower
    pad_x = unit(0.85, "cm"), # adding space between left edge of plot and north arrow, centering underneath scale bar
    pad_y = unit(0.8, "cm") # adding space between top edge of plot and north arrow, ensuring underneath scale bar
  ) +
  coord_sf(
    # cropping the plot to the edge of the data frame
    xlim = c(co_bbox["xmin"] + 0.31, co_bbox["xmax"] - 0.32),
    ylim = c(co_bbox["ymin"] + 0.21, co_bbox["ymax"] - 0.21)
  ) +
  theme_bw() +
  labs(
    # changing labels for x/y axes, title, and legend titles
    x = "Latitude (degrees)",
    y = "Longitude (degrees)",
    title = "Change in Well Water Level Over 40 Years\nin Six Colorado Aquifers",
    color = "Forty Year Difference (ft)",
    fill = "Aquifer"
  ) +
  theme(
    # customizing the plot
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), # centering, sizing, and bolding title
    axis.title = element_text(size = 11, face = "bold"), # sizing and bolding axis titles
    legend.title = element_text(size = 10, face = "bold"), # sizing and bolding legend titles
    axis.text = element_text(size = 9), # sizing axis text
    legend.text = element_text(size = 9) # sizing legend text
  )

```


# References

::: {#refs}
:::