---
title: "Plotting Assignment 6"
subtitle: "Compound Plots and Images"
author: "Haley Brumberger"
date: "`r format(Sys.Date(), '%d %b %Y')`"
number-sections: true
number-offset: 0
toc: true
toc-depth: 2
fig-width: 6
fig-height: 4
df-print: kable
embed-resources: true
format: 
  html: 
    highlight-style: dracula
    code-tools: true
    code-fold: show
    code-summary: "Show the code"
    code-line-numbers: true
    code-link: true # only works if code-line-numbers: false
    toc-float: true
  docx: 
    toc-title: "Table of contents"
    fig-dpi: 600
    execute:
      echo: false
knitr: 
  opts_chunk: 
    fig.path: "plots/"
    fig.keep: "all"
    dev: ['png', 'pdf']
    dev.args: 
      pdf: 
        encoding: 'WinAnsi'
        useDingbats: false
bibliography: references.bib # if using references, need a .bib file
#csl: gpl.csl # find your preferred citation style at https://www.zotero.org/styles
#css: style.css # optional html styling
crossref:
  fig-prefix: Figure
  tbl-prefix: Table
  ref-hyperlink: true
editor: source
---

# Setup

Using `r R.version.string` , @R, and ggplot2 version `r packageVersion("ggplot2")` from the tidyverse [@tidyverse].

```{r}
#| label: setup
#| message: false

# load packages
library(tidyverse)
library(tidyr)
library(stringr)
library(patchwork)
library(cowplot)
library(ggtext)
library(gganimate) # animate plots
library(plotly) # make plots interactive
library(gapminder)
library(magick)
library(scales)

# load script files
source("scripts/plotting_functions.R")
```

# Data

```{r}
# load data from website/CSV

apod <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-20/apod.csv",
  show_col_types = FALSE
)
```

# Plots {#sec-plots}

## Making my own theme
```{r}
my_theme <- theme_bw() + # making my own plot theme
  theme(
    # pretty-ifying the plot by settign line color and width for the panel grid major/minor x/y
    panel.grid.major.x = element_line(color = "black", linewidth = 0.05),
    panel.grid.minor.x = element_line(color = "black", linewidth = 0.025),
    panel.grid.major.y = element_line(color = "black", linewidth = 0.05),
    panel.grid.minor.y = element_line(color = "black", linewidth = 0.025),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12), # adjusting the plot title to be centered, bolded, bigger
    axis.title = element_text(size = 10), # adjusting the axis title to be smaller
    axis.text = element_text(size = 9), # adjusting the axis text to be smallish
    plot.subtitle = (element_text(hjust = 0.5, size = 11)) # centering and making plot subtitle smaller than plot title
  )
```

## Plot 1: Features of Interest
Overall, @fig-1a shows that, of the 5 features I chose (Nebula, Comet, Spiral, Supernova, and Dwarf), Nebula is by far the most mentioned feature across the dataset, which spans from 2007-2025. Comets are mentioned less than 1/2 the amount of times Nebulas are mentioned, and Spirals are mentioned approximately 1/2 the amount of times Comets are mentioned. Supernovas are mentioned fewer than 100 times, and Dwarfs are mentioned fewer than 20 times. Considering that this dataset is ~18 years of data (and 365 * 18 = 6,570), none of these features are actually mentioned that many times; Nebula, my most-mentioned feature at ~725 mentions, is only mentioned in ~11% of all the pictures, while Dwarf (my least mentioned feature) is only mentioned in ~0.2% of all pictures. 
```{r}
#| label: fig-1a
#| fig-cap: Bar plot exploration of five features of interest (Nebula, Comet, Spiral, Supernova, and Dwarf) in NASA's astronomy picture of the day (APOD) dataset from 2007-2025.

# find the pictures that contain 5 astronomy features of interest (nebula, dwarf, comet, spiral [galaxies], supernova)
apod_foi <- apod |>
  mutate(
    feature = str_extract_all(
      title,
      regex("nebula|dwarf|comet|spiral|supernova", ignore_case = TRUE)
    )
  ) |>
  unnest(feature) # unnest the information in the columns

# make a new dataframe, extract year from the date column in apod_foi
apod_foi_counts <- apod_foi |>
  mutate(
    year = date |> year()
  ) |>
  summarize(.by = c("feature", "year"), n = n()) |> # summarize by feature and add an n column for counts
  mutate(
    feature = fct_inorder(factor(feature), ordered = TRUE) # mutate so that feature is an ordered factor so the plot is highest -> lowest # of occurrences
  ) |>
  complete(feature, year, fill = list(n = 0)) # complete the dataset with 0s where there is no data (e.g., adding 0s to blank rows/features that are not documented in certain years)

# plot occurrences of each feature
plot1a <- apod_foi_counts |>
  ggplot() +
  aes(x = feature, y = n, fill = feature) +
  geom_col() +
  scale_y_continuous(
    breaks = seq(0, 800, 100), # setting the y scale major breaks to go from 0-800 in steps of 100
    minor_breaks = seq(0, 800, 20), # setting the y scale minor breaks to go from 0-800 in steps of 25
    expand = expansion(add = c(0, 50)) # adding 100 to the top of the y scale so it's easier to see where the top of the biggest column is; bringing the bottom of the columns all the way down to be flush with the x axis
  ) +
  labs(
    # renaming x, y axes, adding title and subtitle
    x = "Feature",
    y = "Total Number of Pictures",
    title = "NASA's Astronomy Picture of the Day",
    subtitle = "Photos of Features of Interest (2007-2025)"
  ) +
  my_theme + # adding my theme from above
  theme(
    # adjusting the plot so the major and minor x gridlines are gone, since the column plot reads horizontally
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none" # suppressing the fill legend
  )

# plotting the plot
plot1a

```

## Plot 2: Features of Interest Through Time
The animated @fig-1b shows how many pictures different features are mentioned in through time, by year (2007-2025). For the most part, the highest-to-lowest order from @fig-1a stays consistent at Nebula -> Comet -> Spiral -> Supernova -> Dwarf. However, in 2012, Spirals are mentioned more than Comets, which represents pretty much the only inversion of the order. The y scale shows how low the actual mentions are--it only goes from 0 to 50, and none of the features reach 50 (Nebula gets to 49 at max). Given how low the total number of mentions are for each feature across the entire dataset, this low per-year number of mentions is unsurprising.
```{r}
#| label: fig-1b
#| fig-cap: Animated bar plot exploring the number of mentions each of the identified features of interest had per year (2007-2025) of NASA's APOD dataset.
#| dev: png
#| dev.args: null
#| cache: true
#| gganimate: # animation options
#| fpsi: 10 # frames per second, default is 10
#| dpi: 300
#| rewind: true # rewind at the end
#| start_pause: 10 # number of frames to show / pause on
#| end_pause: 10 # number of frames to show / pause on

apod_foi_counts <- apod_foi_counts |>
  mutate(
    year = as.integer(year) # ensure year is a numeric column for ease of interpreting the frames in the animated plot
  )

# make the plot
plot1b <- apod_foi_counts |>
  ggplot() +
  aes(x = feature, y = n, fill = feature) +
  geom_col() +
  theme_bw() +
  scale_y_continuous(
    expand = expansion(add = c(0, 1)), # setting the y axis so that the columns are flush at the bottom with the x axis
    breaks = seq(0, 50, 10), # setting the major y axis breaks from 0-50 in steps of 10
    minor_breaks = seq(0, 50, 2) # setting the minor y axis breaks from 0-50 in steps of 2
  ) +
  labs(
    # renaming x, y axes and adding title and subtitle
    x = "Feature",
    y = "Number of Picture Mentions",
    title = "Number of Picture Mentions per Feature per Year",
    subtitle = "{frame_time}" # put each frame's current year in the subtitle because I couldn't figure out how to make it a label somewhere else without blinking with each frame, this keeps track of what year the data is shown for
  ) +
  my_theme + # adding my theme from above
  theme(
    panel.grid.major.x = element_blank(), # disappearing the x grid
    legend.position = "none", # suppressing the fill legend
  ) +
  transition_time(year) + # setting the transition time to be each year
  ease_aes("linear") # setting the transition between

plot1b # tried to use ggplotly with a slider but the bar/column plot wouldn't work with that; geom_point worked but it wasn't as easy to see the values moving around from year to year, so I decided to just go with gganimate
```

## Plot 3: Favorite Feature Through Time
@fig-2a shows the Percentage of Daily Photos per Year (2007-2025) that mention Supernovas. Of the 5 features of interest, Supernovas were the second-least mentioned overall, so it is unsurprising that this feature repsents <2% of daily mentions for every year. The Supernova mentions peak in 2011 in approximately 1.9% of daily photos, and bottom out in 2010 with mentions in less than 0.3% of daily photos. It is interesting that this pattern goes from the smallest percentage to the largest percentage in a matter of 1 year, but the difference is actually so small (~1.6%, which equates to ~5 photos) that it is basically negligible. It is also interesting that in 2023, 2024, and 2025, Supernovas were mentioned the same percentage of time (in 6 photos, at ~1.6% of daily mentions).
```{r}
#| label: fig-2a
#| fig-cap: Line plot showing the percentage of daily photos with Supernovas mentioned per year (2007-2025). Note how low the y-scale is (from 0 to 2%).
apod_counts_supernova <- apod_foi_counts |>
  filter(feature == "Supernova") |>
  mutate(
    percent_mentioned_per_year = n / 365 # dividing n (number of mentions per year) by 365 to get a rough percentage of days this feature was mentioned per year
  )

plot2a <- apod_counts_supernova |>
  ggplot() +
  aes(x = year, y = percent_mentioned_per_year) +
  geom_point(color = "#47be49") + # setting the color to be similar to the supernova color from the bar plot above
  scale_y_continuous(
    labels = label_percent(),
    breaks = seq(0, 0.03, 0.005),
    minor_breaks = seq(0, 0.03, 0.001),
    expand = expansion(add = c(0.0028, 0.0009))
  ) + # setting y axis major and minor breaks, and expanding the axes to start at 0% and end at 2%
  scale_x_continuous(
    breaks = seq(2007, 2025, 3), # setting x axis major ticks 2007-2025 by steps of 3
    minor_breaks = seq(2007, 2025, 1), # setting x axis minor ticks 2007-2025 by steps of 1
    expand = expansion(add = c(0.5, 0.5)) # setting x axis expansion so 2007 and 2025 are close to the edges of the plot
  ) +
  geom_line(color = "#47be49") + # setting the color to be similar to the supernova color from the bar plot above
  labs(
    x = "Year",
    y = "% of Daily Mentions per Year",
    title = "Percentage of Daily Photos with Supernovas",
    subtitle = "(2007-2025)"
  ) +
  my_theme + # adding my theme from above
  theme(
    legend.position = "none" # suppressing the legend
  )

# plotting the plot
plot2a

```

## Plot 4: Interactive Plot--Favorite Feature Through Time
@fig-2b is an interactive version of @fig-2a. It shows the same data, but you can hover over each point to see the actual percentage values that correspond to each year. 
```{r}
#| label: fig-2b
#| fig-cap: Interactive line plot showing the percentage of daily photos with Supernovas mentioned per year. Note how low the y-scale is (from 0 to 2%).

# making a new dataframe with percentages (*100) for the hover-over popup labels to be accurate in ggplotly without having to multiply by 100 in the plot such that the "* 100" shows up in the label, changing the column labels to show up accurately/nice-looking-ly in the plot
apod_counts_supernova2 <- apod_counts_supernova |>
  mutate(
    `Percentage of Photos (%)` = percent_mentioned_per_year * 100, # making a new column that is the percent mentioned per year in percent form rather than decimal form; naming it the way I want it to show up in the box when you hover over the interactive data point
    `Year` = year # naming the year column the way I want it to show up in the box when you hover over the interactive data point
  )

plot2b <- apod_counts_supernova2 |>
  ggplot() +
  aes(x = `Year`, y = `Percentage of Photos (%)`) +
  geom_point(color = "#47be49") + # setting color as in plot above
  geom_line(color = "#47be49") + # setting color as in plot above
  theme_bw() +
  #scale_y_continuous(labels = scales::label_percent()) # removed percentage scale for y axis since it doesn't translate into ggplotly anyway
  #scale_x_continuous(
  #breaks = seq(2007, 2025, 3), # this doesn't translate into interactive / dynamic ticks ggplot either
  #minor_breaks = seq(2007, 2025, 1) # this doesn't translate into interactive / dynamic ticks ggplot either
  labs(
    # relabeling x, y axes, giving title and subtitle
    x = "Year",
    y = "% of Total Mentions",
    title = "Percentage of Daily Photos with Supernovas",
    subtitle = "(2007-2025)"
  ) +
  theme(
    legend.position = "none" # suppressing legend (just in case)
  )

# ggplotly the plot to make it interactive
plot2b_interactive <- ggplotly(plot2b, dynamicTicks = TRUE) |> # setting dynamic ticks for zooming in/around the plot
  layout(yaxis = list(ticksuffix = "%")) # making the ylabels have % signs

# plotting the interactive ggplotly
plot2b_interactive

```

## Plot 5: Dominant Colors in Favorite Feature Photo
@fig-3 is a bar plot that displays the 5 most common colors shown in my favorite photo of my favorite feature. The explanation of the photo ("Filaments of the Vela Supernova Remnant") from the website is as follows: "The explosion is over, but the consequences continue. About eleven thousand years ago, a star in the constellation of Vela could be seen to explode, creating a strange point of light briefly visible to humans living near the beginning of recorded history. The outer layers of the star crashed into the interstellar medium, driving a shock wave that is still visible today. The featured image captures some of that filamentary and gigantic shock in visible light. As gas flies away from the detonated star, it decays and reacts with the interstellar medium, producing light in many different colors and energy bands. Remaining at the center of the Vela Supernova Remnant is a pulsar, a star as dense as nuclear matter that spins around more than ten times in a single second." As you can see, dark colors dominate this photo, with the most common color being essentially black. There is also then a dark red, then dark blue, dark orage, and finally medium periwinkle.
```{r}
#| label: fig-3
#| fig-cap: Bar plot showing the 5 most common colors shown in NASA's APOD from April 16th, 2024. The photo is titled "Filaments of the Vela Supernova Remnant" and image credit goes to CTIO, NOIRLab, DOE, NSF, AURA; Processing by T. A. Rector (U. Alaska Anchorage), M. Zamani & D. de Martin ('s NOIRLab).

apod_fav <- apod |> # filtering dataset to my fav photo
  filter(date == "2024-04-16")

# reading in the image from the URL
img <- image_read("https://apod.nasa.gov/apod/image/2404/VelaSnr_CTIO_3989.jpg")

# determining the rgb (red, green, blue) composition of the 5 most frequently occurring colors in my image
img_rgb <- img |> image_quantize(max = 5) |> image_data(channels = "rgb")

# making a tibble that holds which of the 5 most frequent colors each pixel is closest to:
colors <- tibble(
  color = rgb(
    img_rgb[1, , ],
    img_rgb[2, , ],
    img_rgb[3, , ],
    maxColorValue = 255
  )
)

# calculating how often each color appears in my image (as percentages)
colors_percent <- colors |>
  summarize(
    .by = "color", # grouping by color
    n = n() # calculating number of appearances
  ) |>
  mutate(
    percentage = (n / sum(n)), # adding a percentage column = number of appearances / total appearances overall
  )


plot3 <- colors_percent |>
  arrange(desc(percentage)) |> # arranging so that the bars go from high -> low
  mutate(color = factor(color, levels = color)) |> # turning color into a factor with levels
  ggplot() +
  aes(x = color, y = percentage, fill = color) +
  geom_col() +
  scale_fill_identity() + # using the color codes in the given data rows
  labs(
    # assigning labels to x, y axes and adding a title and subtitle
    title = "Colors Present in Supernova Photo",
    subtitle = "NASA's Astronomy Picture of the Day: April 16th, 2024",
    x = "Color",
    y = "Percentage of Image"
  ) +
  scale_y_continuous(
    # setting the labels for the y scale to convert to percentages
    labels = label_percent(),
    breaks = seq(0, 0.7, .10), # setting the major breaks at 10%
    minor_breaks = seq(0, 0.7, .02), # setting the minor breaks at 2%
    expand = expansion(add = c(0, 0.05)) # adding 5% to the top of the y scale so it ends at 70% rather that cutting off
  ) +
  my_theme +
  theme(
    panel.grid.major.x = element_blank(), # getting rid of major x gridlines since it's a vertical bar plot
  )

# plotting the plot
plot3

```

## Plot 6: The Image Itself
@fig-4 shows the same bar plot as @fig-3 overlying the photo itself ("Filaments of the Vela Supernova Remnant" from April 16th, 2024). The photo actually appears a lot lighter in the background than I would expect given that a very dark color (essentially black) is more than 65% of the photo and that the 3 next most common colors are also dark. It is surprising that the light oranges and blues that show up in the supernova filaments are not included in this "most common colors" plot.
```{r}
#| label: fig-4
#| fig-cap: Same bar plot visualization as Figure 3 (5 most common colors in my favorite photo), but overlying the actual image/my favorite photo from April 16th, 2024.
#| fig-width: 8
#| fig-height: 8

plot3_for_plot4 <- plot3 +
  geom_col(color = "white") + # setting outline color of columns to be white so they show up on the dark background
  theme(
    panel.background = element_blank(), # getting rid of the panel background so it's transparent
    plot.background = element_blank(), # getting rid of the plot background so it's transparent
    panel.grid.major.x = element_blank(), # suppressing x panel grid lines since it's a vertical column plot
    panel.grid.major.y = element_line(color = "white", linewidth = 0.1), # setting width of the major y panel grid lines
    panel.grid.minor.y = element_line(color = "white", linewidth = 0.05), # setting width of the minor y panel grid lines
    axis.text = element_text(
      color = "white", # turning axis text white so it shows up on dark background
      face = "bold", # making axis text bold so it shows up better on dark background
      size = 11 # setting axis text size
    ),
    axis.title = element_text(
      color = "white", # turning axis title white so it shows up on dark background
      face = "bold", # making axis title bold so it shows up better on dark background
      size = 12 # setting axis title size
    ),
    plot.title = element_text(
      color = "white", # turning plot title white so it shows up on dark background
      face = "bold", # making plot title bold so it shows up better on dark background
      hjust = 0.5, # centering plot title
      size = 16 # making plot title big
    ),
    plot.subtitle = element_text(
      color = "white", # turning plot subtitle white so it shows up on dark background
      face = "bold", # making plot subtitle bold so it shows up better on dark background
      hjust = 0.5, # centering plot subtitle
      size = 15 # making plot subtitle big
    ),
    axis.ticks = element_line(color = "white"), # turning axis ticks white so they show up on dark background
    panel.border = element_rect(color = "white") # setting panel border to be white so it shows up on dark background
  )


plot4 <- # making the plot
  ggdraw() + # setting the canvas
  draw_image(img, x = 0, y = 0, width = 1, height = 1) + # adding the image, setting position and size
  draw_plot(
    # overlaying plot3_for_plot4
    plot3_for_plot4,
    x = 0.05, # setting x position off the left edge
    y = 0.05, # setting y position off the bottom edge
    width = 0.9, # setting the plot width to be a little smaller so it fits squarely in the image
    height = 0.9 # setting the plot height to be a little smaller so it fits squarely in the image
  ) +
  draw_label(
    # adding label (the title of the image)
    "Filaments\nof the Vela Supernova Remnant", # label text string
    color = "white", # label color to show up on dark background
    size = 12, # label text size
    fontface = "bold", # making label bold so it shows up on dark background better
    x = 0.01, # adjusting the x positioning to be close to the left edge
    y = 0.03, # adjusting the y positioning to be a little up from the bottom
    hjust = 0
  )

# plotting the plot
plot4
```

## Plot 7: Compound Plot 1-3
@fig-5 shows @fig-1a (A), @fig-2a (B), and @fig-4 (C) in a compound plot, illustrating features of interest from NASA's APOD, then zooming into Supernova mentions, and finally representing my favorite Supernova photo.
```{r}
#| fig-width: 10
#| fig-height: 12
#| label: fig-5
#| fig-cap: Compound plot showing A) Figure 1 (total mentions of 5 features of interest in NASAs APOD dataset from 2007-2025; B) Figure 2 (percentage of photos that mention Supernovas per year from 2007-2025); and C) Figure 4 (dominant colors in the APOD from April 16th, 2024 of Supernova remnants).

# making a new plot 3 to fit over a new scaled image for plot 5
plot3_for_plot4and5 <- plot3 + # making a new version of plot 3 for plot 4 in the composite plot
  geom_col(color = "white") + # setting the outline to be white
  theme(
    panel.background = element_blank(), # getting rid of panel background so it's transparent
    plot.background = element_blank(), # getting rid of plot background so it's transparent
    panel.grid.major.x = element_blank(), # suppressing x gridlines since it's a vertical bar plot
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "white", linewidth = 0.05), # setting gridline color and widths (also below)
    panel.grid.minor.y = element_line(color = "white", linewidth = 0.025),
    axis.text = element_text(
      # setting axis text specs
      color = "white",
      face = "bold",
    ),
    axis.title = element_text(
      # setting axis title specs
      color = "white",
      face = "bold",
    ),
    plot.title = element_text(
      # setting plot title specs
      color = "white",
      face = "bold",
      hjust = 0.5,
    ),
    plot.subtitle = element_text(
      # setting plot subtitle specs
      color = "white",
      face = "bold",
      hjust = 0.5,
    ),
    axis.ticks = element_line(color = "white"), # setting axis ticks to be white to show up on dark background
    panel.border = element_rect(color = "white") # setting panel border white to show up on dark background
  )

plot4_forplot5 <- # making a new version of plot 4 for the composite plot
  ggplot() + # setting the scale to be within a ggplot coordinate system
  draw_image(
    img, # adding the image
    x = 0, # setting the position of the image
    y = 0,
    scale = 2, # scaling the image so plot 3 fits over it
  ) +
  draw_plot(
    plot3_for_plot4and5, # adding plot3 version above
  ) +
  draw_label(
    # adding a label
    "Filaments of the Vela Supernova Remnant",
    color = "white",
    size = 10,
    fontface = "bold",
    x = 0,
    y = 0.005,
    hjust = 0 # justifying to the bottom left of the plot coordinate system
  )

(plot1a | plot2a & plot_layout(height = 0.5, width = 1, nrow = 1)) / # putting plot1a and plot2a next to each other in one row
  (plot4_forplot5 + plot_layout(height = 6, width = 2)) & # putting plot4 underneath so it stretches across, making it taller for better visibility
  theme(
    # resetting the theme/font size for all the plots
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 13),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12)
  ) &
  plot_annotation(
    tag_levels = "A" # adding labels (A, B, C)
  )

```


# References

::: {#refs}
:::
